Seismic parameter estimation and the Canadian crust

### 2013-06-27 paper/draftBP.tex
paper

  The primary data utilized in this study are computed from teleseismic P-wave seismograms representing discrete seismic events. These events are comprised of seismic records representing more than 700 earthquake sources occuring between the years 2000 and 2012 at 343 broadband seismic stations across Canada. Seismic stations are selected from all available regional and national networks including CNSN, Polaris, FedNor and Chasme. Events are filtered by the epicentral distance from source to receiver with only the events within a 30 to 100 degree window being included. Seismic events are further filtered by hand picking those with reasonable signal to noise ratio and of sufficient impulsiveness that arrival times can be accurately measured. After selection and filtering more than 80,000 events are available for further processing.

@@

\subsection{Vp/Vs method}

@@

  A well tested and widely published method for extracting the ratio, $R=\frac{V_P}{V_S}$, where $V_P$ is P-wave velocity and $V_S$ is S-wave velocity and crustal thickness, or depth to Moho, $H$, is outlined by Zhu and Kanamori [2000], hereafter ZK. This method takes advantage of the differential arrival times between the S-wave reflected phases $Ps$, $PpPs$ $PsPs$, $PpSs$ and $PsSs$ and the direct P-wave arrival $Pp$, Figure ~\ref{fig:reflectedPhases}. Note that $PsPs$ and $PpSs$ have an equal number of $P$ and $S$ phases, and the energy for these two phases arrive simultaneously and are indistinguishable. For a range of slowness values, $p$, the differential arrival times, $t(p)$, trace moveout curves for each phase arrival given by

% Travel time equations
\begin{equation}
t_{Ps}(p_i)=H \left[ \sqrt{ \left(\frac{V_p}{V_s}\right)^2 - p_i^2V_p^2} - \sqrt{1 - p_i^2V_P^2} \right]
\end{equation}

\begin{equation}
t_{Pps}(p_i)=H \left[ \sqrt{ \left(\frac{V_p}{V_s}\right)^2 - p_i^2V_p^2} + \sqrt{1 - p_i^2V_P^2} \right]
\end{equation}

\begin{equation}
t_{Pss}(p_i)= 2H  \sqrt{ \left(\frac{V_p}{V_s}\right)^2 - p_i^2V_p^2}
\end{equation}

@@

@@

\begin{equation}
s(H,R) = \sum_{j=1}^{3} S_j \sum_{i=1}^N w_jr_i(t_j)
\end{equation}

@@

and $S_j$ is the semblance weight for the $j^{th}$ phase, time $t_j$ is calculated from the corrisponding travel time function for a given $H$, $R$ as a function of slowness $p_i$ and $N$ is the number of receiver functions. Multiplying by the semblance weighting sharpens the stacked image, $s(H,R)$, and exhibits better resolution when selecting between stacked models. The model which stacks the most coherent energy is used to provide the best estimates for the bulk crustal parameters $R$ and $H$ under a given seismic station.

  For a given synthetic receiver function
\subsection{Full Gridsearch Method}

  Owing to increasing computational availability it is now reasonably efficient to extend the ZK approach and perform the stack along all three seismic parameters, $V_P$, $R$ (therefore by definition $V_S$) and $H$.

\subsection{Vp data}
   Accompanying the processed estimates outlines above are data from controlled source experiments collected and compiled by external sources. The data was compiled by Mooney (personal communication, 2012).

\subsection{Crust 2.0 data}
   Accompanying the processed estimates outlines above are data from controlled source experiments collected and compiled by external sources. The data was compiled by Mooney (personal communication, 2012).

@@

\subsection{Comparisons}
As regional studies exist which utilize similar methods and have corresponding parameter estimates for some of the stations used in this study it is possible to directly compare values to those previously published. A comprehensive study in the Hudson Bay region of the Canadian Shield (Thomson et. al., 2010) uses an approach similar to the Zhu and Kanamori method published estimates for $H$ and $R=\frac{V_P}{V_S}$ for 35 stations. 5 stations with $R$ estimates above 1.8 are removed due to the uncertainty surrounding these high values. There is strong correlation of 0.95 between crustal thickness values for both datasets (Figure~\ref{fig:thompsonCompH}). The velocity ratio data has a lower correlation of 0.5 (Figure~\ref{fig:thompsonCompR}). [Compare this value to uncertainty / deviation in the data? ].

@@

% AGU does not want a .bib or a .bbl file. Please copy in the contents of your .bbl file here.

\begin{thebibliography}{}

@@

%Reference citation examples:

%...as shown by \textit{Kilby} [2008].
%...as shown by {\textit  {Lewin}} [1976], {\textit  {Carson}} [1986], {\textit  {Bartholdy and Billi}} [2002], and {\textit  {Rinaldi}} [2003].
%...has been shown [\textit{Kilby et al.}, 2008].
%...has been shown [{\textit  {Lewin}}, 1976; {\textit  {Carson}}, 1986; {\textit  {Bartholdy and Billi}}, 2002; {\textit  {Rinaldi}}, 2003].
%...has been shown [e.g., {\textit  {Lewin}}, 1976; {\textit  {Carson}}, 1986; {\textit  {Bartholdy and Billi}}, 2002; {\textit  {Rinaldi}}, 2003].

%...as shown by \citet{jskilby}.
%...as shown by \citet{lewin76}, \citet{carson86}, \citet{bartoldy02}, and \citet{rinaldi03}.
%...has been shown \citep{jskilbye}.
%...has been shown \citep{lewin76,carson86,bartoldy02,rinaldi03}.
%...has been shown \citep [e.g.,][]{lewin76,carson86,bartoldy02,rinaldi03}.
%
% Please use ONLY \citet and \citep for reference citations.
% DO NOT use other cite commands (e.g., \cite, \citeyear, \nocite, \citealp, etc.).

@@

M. G. Bostock, M. R. Kumar, 2010. Bias in seismic estimates of crustal properties, Geophys. J. Int., 182, 403-407.
N. I. Christensen, 1996. Poisson's ratio and crustal seismology, J. Geophys. Res., 101, 3139–3156,
R. J. Durrheim, W. D. Mooney, 1991. Archean and Proterozoic crustal evolution, Geology, 19, 606-609.
W. D. Mooney. 2012. Personal communication. Compiled GSC active source data for the Canada.
D.A. Thompson, I.D. Bastow, G. Helffrich, J-M. Kendall, J. Wookey, D.B. Snyder, D.W. Eaton, 2010. Precambrian crustal evolution: Seismic constraints from the Canadian Shield, Earth and Planetary Science Letters, Volume 297, Issues 3–4, 1 September 2010, Pages 655–666
L. Zhu, H. Kanamori, 2000. Moho depth variation in Southern California from teleseismic receiver functions, J. Geophys Res, 105, 2969-2980.
G. Zandt, C. J. Ammon, 1995. Continental crust composition constrained by measurements of crustal Poisson's ratio, Nature, 374, 152-154.

%
%  SECTION HEADS
%
%% ------------------------------------------------------------------------ %%

ABSTRACT
   It has been suggested that processes driving crustal formation in the Archean and Proterozoic were dissimilar and produced crusts with unique bulk properties and average thicknesses. Existing models based on fractionating mantle composition or evolving mantle convection require accurate estimates of the geological and geophysical properties of crustal provinces to better understand the details of early continental formation (R. Durrheim, W. Mooney, 1991). Fifteen years of publicly accessible teleseismic data from all available Canadian seismic stations are binned in horizontal slowness and deconvolved into receiver functions. We apply a new stacking method (Bostock and Kumar, 2010) to retrieve estimates of bulk crustal velocities (Vp, Vs) and thickness H from these data under the assumption of 1-D structure. Bootstrap error analysis is performed for each station dataset and results are compared to the results produced from alternate methods (Zhu and Kanamori, 2000). Cross-analyzing these results with the mineral and rock seismic property database of Christensen (1996) will afford improved constraints on bulk geological composition of the Canadian landmass. These results will be used to evaluate competing models of early crustal formation.

% Capitalize the first letter of each word (except for
% prepositions, conjunctions, and articles that are
% three or fewer letters).

% AGU follows standard outline style; therefore, there cannot be a section 1 without
% a section 2, or a section 2.3.1 without a section 2.3.2.
% Please make sure your section numbers are balanced.
% ---------------
% Level 1 head
%
% Use the \section{} command to identify level 1 heads;
% type the appropriate head wording between the curly
% brackets, as shown below.
%
%An example:
%\section{Level 1 Head: Introduction}
%
% ---------------
% Level 2 head
%
% Use the \subsection{} command to identify level 2 heads.
%An example:
%\subsection{Level 2 Head}
%
% ---------------
% Level 3 head
%
% Use the \subsubsection{} command to identify level 3 heads
%An example:
%\subsubsection{Level 3 Head}
%
%---------------
% Level 4 head
%
% Use the \subsubsubsection{} command to identify level 3 heads
% An example:
%\subsubsubsection{Level 4 Head} An example.
%
%% ------------------------------------------------------------------------ %%
%
%  IN-TEXT LISTS
%
%% ------------------------------------------------------------------------ %%
%
% Do not use bulleted lists; enumerated lists are okay.
% \begin{enumerate}
% \item
% \item
% \item
% \end{enumerate}
%
%% ------------------------------------------------------------------------ %%
%
%  EQUATIONS
%
%% ------------------------------------------------------------------------ %%

% Single-line equations are centered.
% Equation arrays will appear left-aligned.

Math coded inside display math mode \[ ...\]
 will not be numbered, e.g.,:
 \[ x^2=y^2 + z^2\]

 Math coded inside \begin{equation} and \end{equation} will
 be automatically numbered, e.g.,:
 \begin{equation}
 x^2=y^2 + z^2
 \end{equation}

% IF YOU HAVE MULTI-LINE EQUATIONS, PLEASE
% BREAK THE EQUATIONS INTO TWO OR MORE LINES
% OF SINGLE COLUMN WIDTH (20 pc, 8.3 cm)
% using double backslashes (\\).

% To create multiline equations, use the
% \begin{eqnarray} and \end{eqnarray} environment
% as demonstrated below.
\begin{eqnarray}
  x_{1} & = & (x - x_{0}) \cos \Theta \nonumber \\
        && + (y - y_{0}) \sin \Theta  \nonumber \\
  y_{1} & = & -(x - x_{0}) \sin \Theta \nonumber \\
        && + (y - y_{0}) \cos \Theta.
\end{eqnarray}

%If you don't want an equation number, use the star form:
%\begin{eqnarray*}...\end{eqnarray*}

% Break each line at a sign of operation
% (+, -, etc.) if possible, with the sign of operation
% on the new line.

% Indent second and subsequent lines to align with
% the first character following the equal sign on the
% first line.

% Use an \hspace{} command to insert horizontal space
% into your equation if necessary. Place an appropriate
% unit of measure between the curly braces, e.g.
% \hspace{1in}; you may have to experiment to achieve
% the correct amount of space.

%% ------------------------------------------------------------------------ %%
%
%  EQUATION NUMBERING: COUNTER
%
%% ------------------------------------------------------------------------ %%

% You may change equation numbering by resetting
% the equation counter or by explicitly numbering
% an equation.

% To explicitly number an equation, type \eqnum{}
% (with the desired number between the brackets)
% after the \begin{equation} or \begin{eqnarray}
% command.  The \eqnum{} command will affect only
% the equation it appears with; LaTeX will number
% any equations appearing later in the manuscript
% according to the equation counter.
%

% If you have a multiline equation that needs only
% one equation number, use a \nonumber command in
% front of the double backslashes (\\) as shown in
% the multiline equation above.

%% ------------------------------------------------------------------------ %%
%
%  LANDSCAPE/SIDEWAYS FIGURE AND TABLE EXAMPLES
%
%% ------------------------------------------------------------------------ %%
%
% For figures, add \usepackage{lscape} to the file and the landscape.sty style file
% to the paper folder.
%
% \begin{figure*}[p]
% \begin{landscapefigure*}
% Illustration here.
% \caption{caption here}
% \end{landscapefigure*}
% \end{figure*}
%
% For tables, add \usepackage{rotating} to the paper and add the rotating.sty file to the folder.
%
% AGU prefers the use of {sidewaystable} over {landscapetable} as it causes fewer problems.
%
% \begin{sidewaystable}
% \caption{}
% \begin{tabular}
% Table layout here.
% \end{tabular}
% \end{sidewaystable}
%
%

>>>>

  The primary data utilized in this study are computed from teleseismic P-wave seismograms representing discrete seismic events. These events are comprised of seismic records representing more than 700 earthquake sources occurring between the years 2000 and 2012 at 343 broadband seismic stations across Canada. Seismic stations are selected from all available regional and national networks including CNSN, Polaris, FedNor and Chasme. Events are filtered by the epicentral distance from source to receiver with only the events within a 30 to 100 degree window being included. Seismic events are further filtered by hand picking those with reasonable signal to noise ratio and of sufficient impulsiveness that arrival times can be accurately measured. After selection and filtering more than 80,000 events are available for further processing.

@@

\subsection{Vp/Vs method} \label{section:VpVsMethod}

@@

  A well tested and widely published method for extracting the ratio, $R=\frac{V_P}{V_S}$, where $V_P$ is P-wave velocity and $V_S$ is S-wave velocity and crustal thickness, or depth to Moho, $H$, is outlined by Zhu and Kanamori [2000], hereafter ZK. This method takes advantage of the differential arrival times between the S-wave reflected phases $Ps$, $PpPs$ $PsPs$, $PpSs$ and $PsSs$ and the direct P-wave arrival $Pp$ (Fig \ref{fig:reflectedPhases}). Note that $PsPs$ and $PpSs$ have an equal number of $P$ and $S$ phases, and the energy for these two phases arrive simultaneously and are indistinguishable. For a range of slowness values, $p$, the differential arrival times, $t(p)$, trace moveout curves for each phase arrival given by

% Travel time equations
\begin{equation} \label{eq:tps}
t_{Ps}(p_i)=H \left[ \sqrt{ \left(\frac{R}{V_P}\right)^2 - p_i^2} - \sqrt{\frac{1}{V_P^2} - p_i^2} \right]
\end{equation}

\begin{equation}
t_{Pps}(p_i)=H \left[ \sqrt{ \left(\frac{R}{V_P}\right)^2 - p_i^2} + \sqrt{\frac{1}{V_P^2} - p_i^2} \right]
\end{equation}

\begin{equation}
t_{Pss}(p_i)= 2H  \sqrt{ \left(\frac{R}{V_P}\right)^2 - p_i^2}
\end{equation}

@@

\begin{figure}
  \centering
    \includegraphics[width=\textwidth]{kinematicRvsH}
  \caption{Kinematic curves traced out by the stacking function, $s(H,R)$, corrisponding to the mapping of fixed travel times $t_{Ps}$, $t_{PpPs}$ and $t_{PsPs}$ into $H$ and $R$ space.}
  \label{fig:kinematicRvsH}
\end{figure}

@@

\begin{equation}  \label{eq:stack}
s(H,R) = \sum_{j=1}^{3} S_j \sum_{i=1}^N w_jr_i(t_j)
\end{equation}

@@

and $S_j$ is the semblance weight for the $j^{th}$ phase, time $t_j$ is calculated from the corrisponding travel time function for a given $H$, $R$ as a function of slowness $p_i$ and $N$ is the number of receiver functions. Multiplying by the semblance weighting sharpens the stacked image, $s(H,R)$, and exhibits better resolution when selecting between stacked models. The function $s(H,R)$ can be thought of as a transformation that maps the $Ps$, $PpPs$ and $PpSs+PsPs$ pulses of the receiver function, $r(t)$,  into $R$ and $H$ space as positive bands, via Eq. \ref{eq:stack}. Constructive interference where these bands intersect will produce a maximum in the stacking function (Fig \ref{fig:kinematicRvsH}). The model, $R$ and $H$, which produce this maximum provides the best estimate for the bulk crustal parameters for a given seismic station. The elongate shape about the point of intersection, with the long axis tilted mostly in the $R$ coordinate, conveys the inherent trade-off between $R$ and $H$ and the lower resolution exhibited by the seismic velocity ratio $R$.

\subsection{Full Gridsearch Method}

  Owing to increasing computational availability it is now reasonably efficient to extend the ZK approach and perform the stack along a range of all three seismic parameters, $V_P$, $R$ (therefore by definition $V_S$) and $H$. This has the advantage of not requiring an assumption on $V_P$ and should produce more accurate estimates, especially above crustal units with anomalous $V_P$. The inherent trade-off between $V_P$ and $R$ is limited, offering reasonable resolution between the two parameters. However, a significant trade-off between $V_P$ and $H$, which is a source of significant correlated error, requires that recovery of $V_P$ should be attempted only in cases where high frequency and high signal-to-noise ratio signals are present in the receiver function data. [[ Could go into more depth here, show the kinematic curves and poor trade-off between $V_P$ and $R$, Figure showing correlation between $V_P$ and $H$]]

  These trade-offs, along with other sources of error such as low signal-to-noise signals resulting from complex geology, are quantified by bootstrap resampling [Efron and Tibshirani, 1986]. Estimates are reprocessed using a random selection of receiver functions, with replacement, 1024 times. An estimate for the error is computed by taking the standard deviation from these trial runs.

\subsection{Additional data}
   Accompanying the processed estimates are data from controlled source experiments collected and compiled from GSC (Geological Survey of Canada) references by Walter Mooney (personal communication, 2012). This data provides $V_P$ and a few $V_S$ estimates for hundreds of experiment locations across Canada, many along traversals across geological provincial boundaries, faults and discontinuities. Some of these data are located within close enough proximity to seismic stations being utilized in this study that they may be used to compare with the $V_P$ and $V_S$ estimates resulting from the full parameter gridsearch.

  Whereas the active source data provide high resolution values at specific sites, the Crust 2.0 model, a global dataset widely cited in academic literature, provides complete coverage across Canada with an attendant trade-off in resolution. The data includes estimates of $V_P$ and $V_S$ for $2^o$ longitude and latitude cells as well as information regarding sedimentary thickness and crustal depth. While Crust 2.0 source data is a superset of the aforementioned active source data, the crustal model is also composed of statistically aggregated data from geographically distant but geologically similar regions. It therefore forms a unique dataset well suited for comparison with the crustal estimates calculated in this study.

@@

\subsection{Comparisons}

  Several published regional studies utilizing the methods outlined in section \ref{section:VpVsMethod} provide parameter estimates for $R$ and $H$ for regionally grouped subsets of Canadian seismic stations.

As regional studies exist which utilize similar methods and have corresponding parameter estimates for some of the stations used in this study it is possible to directly compare values to those previously published. A comprehensive study in the Hudson Bay region of the Canadian Shield (Thomson et. al., 2010) uses an approach similar to the Zhu and Kanamori method published estimates for $H$ and $R=\frac{V_P}{V_S}$ for 35 stations. 5 stations with $R$ estimates above 1.8 are removed due to the uncertainty surrounding these high values. There is strong correlation of 0.95 between crustal thickness values for both datasets (Figure~\ref{fig:thompsonCompH}). The velocity ratio data has a lower correlation of 0.5 (Figure~\ref{fig:thompsonCompR}). [Compare this value to uncertainty / deviation in the data? ].

@@

\begin{thebibliography}
Bostock, M. G. (1998), Mantle stratigraphy and the evolution of the Slave province, J. Geophys. Res., 103, 21,183-21,200.

Bostock, M. G., M. R. Kumar (2010), Bias in seismic estimates of crustal properties, J. Geophys. Int., 182, 403-407.

Christensen, N. I. (1996), Poisson's ratio and crustal seismology, J. Geophys. Res., 101, 3139–3156.

Darbyshire, F. A., D. W. Eaton, A. W. Frederiksen, E. Leila (2006), New insights into the lithosphere beneath the Superior Province from Rayleigh wave dispersion and receiver function analysis, J. Geophys. Int., 169, 1043-1068.

Durrheim, R. J., W. D. Mooney (1991), Archean and Proterozoic crustal evolution, Geology, 19, 606-609.

Eaton, D. W., S. Dineva, R. Mereu (2005), Crustal thickness and Vp/Vs variations in the Grenville orogen (Ontario, Canada) from analysis of teleseismic receiver functions, Tectonophysics, 420, 223-238.

Efron, B., R. Tibshirani (1986), Bootstrap Methods for Standard Errors, Confidence Intervals, and Other Measures of Statistical Accuracy, Statistical Science, 1, 54-75.

Golub, G. H., M. Heath, G. Wahba (1979), Generalized cross-validation as a method for choosing a good ridge parameter, Technometrics, 21, 215-223.
Mooney, W. D. (2012), Personal communication. Compiled GSC active source data for the Canada.

Perry, H. K. C., D. W. S. Eaton, A. M. Forte (2002) LITH5.0: a revised crustal model for Canada based on Lithoprobe results,  J. Geophys. Int., 150, 285-294.

Thompson, D. A., I. D. Bastow, G. Helffrich, J.-M. Kendall, J. Wookey, D. B. Snyder, D. W. Eaton (2010), Precambrian crustal evolution: Seismic constraints from the Canadian Shield, Earth and Planetary Science Letters, 297, 655–666.

Zhu, L., H. Kanamori (2000), Moho depth variation in Southern California from teleseismic receiver functions, J. Geophys. Res., 105, 2969-2980.

Zandt, G., C. J. Ammon (1995), Continental crust composition constrained by measurements of crustal Poisson's ratio, Nature, 374, 152-154.

@@

@@
